Post-marital residence rules and transmission pathways in cultural hitchhiking

### 2024-07-13 manuscript/main_revision_changes.tex
Adding paper source file

% Please add a significance statement to explain the relevance of your work
\significancestatement{Population dispersal and admixture can promote the spread of adaptive cultural traits, such as agriculture, in a geographic region. 
In some circumstances, this process can lead to the spread of selectively neutral traits that are transmitted together with adaptive ones, leading to what some refer to as \textit{cultural hitchhiking}. 
Here, through an individual-based simulation, we demonstrate that factors such as post-marital residence rules, probability of post-marital social learning, sex-biased transmission, and transmission pathways can all impact the degree of hitchhiking and inter-group cultural diversity. 
Our results demonstrate how intergroup exogamy is a complex phenomenon that can theoretically lead to a substantially wide range of expected patterns. }

@@

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Cultural hitchhiking | Transition to farming | Pathways of transmission | Post-marital residence rules | Individual-based modelling}

\begin{abstract}
Cultural evolutionary processes can often lead to a statistical association between neutral and adaptive traits during episodes of population dispersal and the introduction of a beneficial technology in a geographic region. 
Here, we examine such cultural hitchhiking processes using an individual-based model that portrays the cultural interaction between a migrant and an incumbent population. 
Our model is loosely based on the interaction between farming and foraging populations during the initial stages of the adoption and diffusion of agricultural practices.
The two populations are characterised by different variants for their neutral and adaptive cultural traits, with the latter set providing a reproductive advantage for the migrant communities over the incumbent ones. 
We explore how the neutral traits of the migrant population spread and how this process is conditioned by the following factors: 1) the possibility of transmission of the adaptive traits; 2) the extent of the increased reproductive advantage provided by the adaptive variants of the migrant population; 3) post-marital residence rules; and 4) how and when neutral traits are transmitted. 
Our results reveal a diverse range of outputs, highlighting the relevance of factors such as the nature of post-martial resocialisation and the specific combination of post-marital residence rules and sex-biased transmission. 
\end{abstract}

@@

In this paper, we focus on the processes involved in the spread of farming and the resulting interactions between migrant agricultural groups and incumbent foraging populations that have occurred in many parts of the world at different times. 

@@

Their dynamics on the spread of farming can be generalised to a wider range of processes involving the interaction between two populations with different adaptive trait variants.

@@

While the movement of individuals across cultural groups can be the result of a wide range of processes, in relatively small-scale societies, such as those associated with the transition to farming, the most common form is inter-group exogamy, i.e. the social practice of marrying outside one's local group.
These episodes of post-marital movement offer the most prominent pathway for cultural transmission across communities.

@@

However, theoretical expectations of what we should observe when specific assumptions and conditions are met remain a comparatively less researched area of investigation. Here we aim to address this issue, focusing in particular on the interplay between post-marital residence rules and resocialisation involving two populations with initially distinct adaptive variants. 
Our goal is to explicitly explore how these processes may lead to cultural hitchhiking and influence the spread of adaptive and non-adaptive cultural trait variants, therefore providing new insights into the diffusion of agriculture.

@@

The spread of agriculture was also characterised by the transmission of neutral cultural traits, such as pottery decoration styles, that were transmitted along with selectively advantageous traits leading to a statistical association between neutral and adaptive traits. 
This process is often referred to as \emph{cultural hitchhiking} (\textit{sensu} \citep{ackland_cultural_2007}), and 
quantitative models of such a phenomenon, or more broadly cultural linkage, have been examined in the past~\citep{ackland_cultural_2007,premo_hitchhikers_2012,yeh_cultural_2019,pardo-gordo_cultural_2021}. 
For example, Ackland et al.~\citep{ackland_cultural_2007} argue that in the presence of a heterogeneous landscape, we should expect to observe a diffusion boundary where the wave of demographic advance is temporarily halted and adaptive trait variants are transmitted to incumbent communities without the hitchhiking cultural variants, leading to the emergence of a permanent `cultural boundary'. 
While processes leading to an association between different cultural traits can vary, from simple functional or cognitive associations to the consequences of certain types of transmission bias (e.g. prestige-biased cultural transmission leading to the adoption of multiple independent traits possessed by the same prestigious individual), the extent by which neutral and adaptive traits can be statistically associated as a consequence of inter-group exogamy remains virtually unexplored.

Here, via an individual-based simulation, we examine how the link between variants of adaptive and neutral traits is maintained or lost due to different post-marital residence rules and transmission pathways within a two-population model. 
Our objective is to provide potential insights into the baseline conditions that can promote or hinder cultural hitchhiking during the spread of beneficial traits between initially distinct human groups.
The analyses we undertake extend from previous models of cultural hitchhiking. For example, Ackland et al.~\citep{ackland_cultural_2007} assume that inter-group transmission occurs only for adaptive traits, while here, we allow both neutral and adaptive traits to be transmitted. 
Our simulation exercise also differs from other demic-cultural diffusion models~\citep[e.g.][]{fort_synthesis_2012}, where neither the interplay between post-marital residence rules and transmission pathways nor the statistical association between neutral and adaptive traits are explored.
Our model does present some similarities to cultural evolutionary models on ethnic markers (e.g.~\citep{boyd_evolution-ethincethnic-markers_1987,mcelreath_share-norm_2003}). 
However, we note that in our study, neutral traits are entirely non-functional, have no signalling value, and dynamics are not driven by cooperation/coordination games. 
In particular, as noted in the model description below, we assume that the association between the adaptive and neutral traits is not the result of the latter promoting indirectly cooperative/coordinating tasks (i.e. acting as a marker trait) but purely the consequences of different patterns of post-marital migration rules and transmission rules.

@@

We consider a population of individuals distributed across several communities, $j$, located at positions $(x_j,y_j)$ in a two-dimensional $10\times10$ grid. 
Each community consists of individuals with the following state variables: \textit{sex}, \textit{age}, and the adopted variants of $z$ neutral traits, $c_1,c_2,\ldots,c_z$ as well as the variants of three adaptive traits, $a_1,a_2,a_3$.
Fig.~\ref{fig:diagram} shows a generalised schematic of the model highlighting its core processes, involving post-marital movement (step 2) and up to two episodes of cultural transmission, namely vertical transmission in the community of origin (step 4) and resocialisation after post-marital migration (step 3).

@@

\begin{figure*}[htb]
\includegraphics[width=17.0cm]{Figs/Figure1.png}
\caption{Generalised schematic of the model with two communities (1). In (2), individuals of the same sex D, E, and J migrate and find a partner in their new communities. In (3), individual E engages into resocialisation via oblique cultural transmission. In the last step (4), the new pairs reproduce giving birth to individual L, M, and N who adopt the cultural variants of one of the parents via vertical transmission.} 
\label{fig:diagram}       
\end{figure*}

@@

Our experiment design is set to explore the specific conditions under which the association between adaptive and neutral variants observed initially in the type B communities persist over time. 
More specifically, we explore how potential differences in growth rates between populations (i.e. different values for $f_1$, $f_2$, and $f_3$) and the degree by which adaptive traits can be adopted (i.e. different values of $\beta$) condition the joint spread (or lack thereof) of the variants of the neutral and adaptive traits of the migrant population ($a_i=1$, $i=1,\ldots,3$ and $c_i=1$, $i=1,\ldots,z$). 
For that, we consider three values for the growth benefits $f_i$ characterising situations of no growth advantage ($f_i=0$, i.e. no difference in the growth rate between both community types), small growth advantage ($f_i=0.005$, i.e. the effective growth rate of type B communities is slightly higher than of type A communities and high growth advantage ($f_i=0.015$, i.e. the effective growth rate of type B communities has almost doubled --- it increased to roughly 0.015 compared to roughly 0.008 for type A communities  (see Fig.~S1, left panel, in the supplementary material)). For the strength of adoption of the adaptive traits $a_i$ we consider two values. While $\beta=-10$ describes the situation of no adoption --- \eqref{eq:adoption} equals almost 0 for all values of numbers of migrants and community sizes ---, $\beta=0$ describes the situation of relatively strong adoption. Fig.~S4 in the supplementary material shows that even for one migrants the adoption probability (per time step) has a value of roughly 0.01.   
Further, we explore  how different transmission pathways (Table~\ref{tab:transCombinations}), biases (controlled by $p_\text{bias}$) and post-marital residence rules (controlled by $p_\text{location}$) interact with the joint spread process.

@@

\section*{Results}

@@

Despite its complexity, our model is far from an exhaustive exploration of the interaction between incumbent and immigrating communities and their effects on the distribution of adaptive and neutral traits. 
For example, our model has assumed that adaptive traits are effectively transmitted without any bias, making the probability of incumbent communities adopting the more beneficial variant from the migrant communities the same as that of the migrant communities switching to the less beneficial variant of the incumbent groups. 
A less conservative scenario for example, one involving a payoff-biased transmission for the adaptive traits where the probability of adopting a more beneficial variant is higher, would have most likely led to a stronger hitchhiking signature. 
It is worth noting that several assumptions and model settings relevant specifically to the spread of agriculture may not apply to other scenarios. 
Our model assumes that migration between populations is the result of inter-group exogamy. We do not consider movements dictated by other factors that might be similarly sex-biased or entail migrations of entire families.
%Similarly, our model envisages interactions occurring when migrant communities are not at carrying capacity. 
One could envisage other contexts in which such an assumption, or the initial relative frequencies of the two populations, is different from the settings we explored.
Further, we assumed that adaptive traits evolve at the group level to portray the nature of a beneficial trait that requires cooperative networks. 
We modelled the transmission of such adaptive traits \textit{as if} communities were individuals, with a frequency-dependent probability of adoption. More complex models, for example inspired by threshold models of collective behaviour~\cite{granovetter_threshold_1978}, could potentially reveal further insights on the coevolution of adaptive and neutral traits. For simplicity, here we explored a simpler scenario that nonetheless captures some specific aspects of subsistence strategies such as farming. 
We note that with our focus on understanding the spread of farming, the analysis of the first phase, i.e. before the population has reached its carrying capacity, is of particular interest. It is widely accepted that farmers usually migrate to new areas in small numbers and experience some sort of competitive growth advantage compared to the incumbent hunter-gatherer communities. If a hitchhiking signature has not been established in this first phase, it is  unlikely to happen in later phases. Naturally, a precise analysis of later phases of the spread dynamic may involve the inclusion of population regulation which is outside the scope of this paper.

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Population dispersal and admixture can promote the spread of adaptive cultural traits, {\color{red}such as agriculture}, in a geographic region. 
In some circumstances, this process can lead to the spread of selectively neutral traits that are transmitted together with adaptive ones, leading to what some refer to as \textit{cultural hitchhiking}. 
Here, through an individual-based simulation, we demonstrate that factors such as post-marital residence rules, probability of post-marital social learning, sex-biased transmission, and transmission pathways can all impact the degree of hitchhiking and inter-group cultural diversity. 
Our results demonstrate how intergroup exogamy is a complex phenomenon that can theoretically lead to a substantially wide range of expected patterns. }

@@

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Cultural Hitchhiking | Transition to farming | Pathways of Transmission | Post-marital residence rules | Individual-based modelling}

\begin{abstract}
Cultural evolutionary processes can often lead to a statistical association between neutral and adaptive traits during episodes of population dispersal and the introduction of a beneficial technology in a geographic region. 
Here, we examine such cultural hitchhiking processes using an individual-based model that portrays the cultural interaction between a migrant and an incumbent population. 
{\color{red}Our model is loosely based on the interaction between farming and foraging populations during the initial stages of the adoption and diffusion of agricultural practices.}
The two populations are characterised by different variants for their neutral and adaptive cultural traits, with the latter set providing a reproductive advantage for the migrant communities over the incumbent ones. 
We explore how the neutral traits of the migrant population spread and how this process is conditioned by the following factors: 1) the possibility of transmission of the adaptive traits; 2) the extent of the increased reproductive advantage provided by the adaptive variants of the migrant population; 3) post-marital residence rules; and 4) how and when neutral traits are transmitted. 
Our results reveal a diverse range of outputs, highlighting the relevance of factors such as the nature of post-martial resocialisation and the specific combination of post-marital residence rules and sex-biased transmission. 
\end{abstract}

@@

{\color{red} In this paper, we focus on the processes involved in the spread of farming and the resulting interactions between migrant agricultural groups and incumbent foraging populations that have occurred in many parts of the world at different times. 

@@

Their dynamics on the spread of farming can be generalised to a wider range of processes involving the interaction between two populations with different adaptive trait variants.}

@@

{\color{red}While the movement of individuals across cultural groups can be the result of a wide range of processes, in relatively small-scale societies, such as those associated with the transition to farming, the most common form is inter-group exogamy, i.e. the social practice of marrying outside one's local group.
These episodes of post-marital movement offer the most prominent pathway for cultural transmission across communities.}

@@

{\color{red}However, theoretical expectations of what we should observe when specific assumptions and conditions are met remain a comparatively less researched area of investigation. Here we aim to address this issue, focusing in particular on the interplay between post-marital residence rules and resocialisation involving two populations with initially distinct adaptive variants. 
Our goal is to explicitly explore how these processes may lead to cultural hitchhiking and influence the spread of adaptive and non-adaptive cultural trait variants, therefore providing new insights into the diffusion of agriculture.}

@@

The spread of agriculture was also characterised by the transmission of neutral cultural traits, such as pottery decoration styles, that were transmitted along with selectively advantageous traits leading to a statistical association between neutral and adaptive traits. 
This process is often referred to as \emph{cultural hitchhiking} (\textit{sensu} \citep{ackland_cultural_2007}), and 
quantitative models of such a phenomenon, or more broadly cultural linkage, have been examined in the past~\citep{ackland_cultural_2007,premo_hitchhikers_2012,yeh_cultural_2019,pardo-gordo_cultural_2021}. 
{\color{red}For example, Ackland et al.~\citep{ackland_cultural_2007} argue that in the presence of a heterogeneous landscape, we should expect to observe a diffusion boundary where the wave of demographic advance is temporarily halted and adaptive trait variants are transmitted to incumbent communities without the hitchhiking cultural variants, leading to the emergence of a permanent `cultural boundary'. 
While processes leading to an association between different cultural traits can vary, from simple functional or cognitive associations to the consequences of certain types of transmission bias (e.g. prestige-biased cultural transmission leading to the adoption of multiple independent traits possessed by the same prestigious individual), the extent by which neutral and adaptive traits can be statistically associated as a consequence of inter-group exogamy remains virtually unexplored.}

Here, via an individual-based simulation, we examine how the link between variants of adaptive and neutral traits is maintained or lost due to different post-marital residence rules and transmission pathways within a two-population model. 
Our objective is to provide potential insights into the baseline conditions that can promote or hinder cultural hitchhiking during the spread of beneficial traits between initially distinct human groups.
{\color{red}The analyses we undertake extend from previous models of cultural hitchhiking. For example, Ackland et al.~\citep{ackland_cultural_2007} assume that inter-group transmission occurs only for adaptive traits, while here, we allow both neutral and adaptive traits to be transmitted. 
Our simulation exercise also differs from other demic-cultural diffusion models~\citep[e.g.][]{fort_synthesis_2012}, where neither the interplay between post-marital residence rules and transmission pathways nor the statistical association between neutral and adaptive traits are explored.
Our model does present some similarities to cultural evolutionary models on ethnic markers (e.g.~\citep{boyd_evolution-ethincethnic-markers_1987,mcelreath_share-norm_2003}). 
However, we note that in our study, neutral traits are entirely non-functional, have no signalling value, and dynamics are not driven by cooperation/coordination games. 
In particular, as noted in the model description below, we assume that the association between the adaptive and neutral traits is not the result of the latter promoting indirectly cooperative/coordinating tasks (i.e. acting as a marker trait) but purely the consequences of different patterns of post-marital migration rules and transmission rules. }

@@

We consider a population of individuals distributed across several communities, $j$, located at positions $(x_j,y_j)$ in a two-dimensional $10\times10$ grid. 
Each community consists of individuals with the following state variables: \textit{sex}, \textit{age}, and the adopted variants of $z$ neutral traits, $c_1,c_2,\ldots,c_z$ as well as the variants of three adaptive traits, $a_1,a_2,a_3$.
{\color{red}Fig.~\ref{fig:diagram} shows a generalised schematic of the model highlighting its core processes, involving post-marital movement (step 2) and up to two episodes of cultural transmission, namely vertical transmission in the community of origin (step 4) and resocialisation after post-marital migration (step 3).}

@@

\begin{figure*}[htb]
\includegraphics[width=17.0cm]{Figs/Figure1.png}
\caption{{\color{red}Generalised schematic of the model with two communities (1). In (2), individuals of the same sex D, E, and J migrate and find a partner in their new communities. In (3), individual E engages into resocialisation via oblique cultural transmission. In the last step (4), the new pairs reproduce giving birth to individual L, M, and N who adopt the cultural variants of one of the parents via vertical transmission.}} 
\label{fig:diagram}       
\end{figure*}

@@

Our experiment design is set to explore the specific conditions under which the association between adaptive and neutral variants observed initially in the type B communities persist over time. 
More specifically, we explore how potential differences in growth rates between populations (i.e. different values for $f_1$, $f_2$, and $f_3$) and the degree by which adaptive traits can be adopted (i.e. different values of $\beta$) condition the joint spread (or lack thereof) of the variants of the neutral and adaptive traits of the migrant population ($a_i=1$, $i=1,\ldots,3$ and $c_i=1$, $i=1,\ldots,z$). 
{\color{red}For that, we consider three values for the growth benefits $f_i$ characterising situations of no growth advantage ($f_i=0$, i.e. no difference in the growth rate between both community types), small growth advantage ($f_i=0.005$, i.e. the effective growth rate of type B communities is slightly higher than of type A communities and high growth advantage ($f_i=0.015$, i.e. the effective growth rate of type B communities has almost doubled --- it increased to roughly 0.015 compared to roughly 0.008 for type A communities  (see Fig.~S1, left panel, in the supplementary material)). For the strength of adoption of the adaptive traits $a_i$ we consider two values. While $\beta=-10$ describes the situation of no adoption --- \eqref{eq:adoption} equals almost 0 for all values of numbers of migrants and community sizes ---, $\beta=0$ describes the situation of relatively strong adoption. Fig.~S4 in the supplementary material shows that even for one migrants the adoption probability (per time step) has a value of roughly 0.01. }  
Further, we explore  how different transmission pathways (Table~\ref{tab:transCombinations}), biases (controlled by $p_\text{bias}$) and post-marital residence rules (controlled by $p_\text{location}$) interact with the joint spread process.

@@

\section*{{\color{red}Results}}

@@

Despite its complexity, our model is far from an exhaustive exploration of the interaction between incumbent and immigrating communities and their effects on the distribution of adaptive and neutral traits. 
For example, our model has assumed that adaptive traits are effectively transmitted without any bias, making the probability of incumbent communities adopting the more beneficial variant from the migrant communities the same as that of the migrant communities switching to the less beneficial variant of the incumbent groups. 
A less conservative scenario for example, one involving a payoff-biased transmission for the adaptive traits where the probability of adopting a more beneficial variant is higher, would have most likely led to a stronger hitchhiking signature. 
{\color{red} It is worth noting that several assumptions and model settings relevant specifically to the spread of agriculture may not apply to other scenarios. 
Our model assumes that migration between populations is the result of inter-group exogamy. We do not consider movements dictated by other factors that might be similarly sex-biased or entail migrations of entire families.
%Similarly, our model envisages interactions occurring when migrant communities are not at carrying capacity. 
One could envisage other contexts in which such an assumption, or the initial relative frequencies of the two populations, is different from the settings we explored.
Further, we assumed that adaptive traits evolve at the group level to portray the nature of a beneficial trait that requires cooperative networks. 
We modelled the transmission of such adaptive traits \textit{as if} communities were individuals, with a frequency-dependent probability of adoption. More complex models, for example inspired by threshold models of collective behaviour~\cite{granovetter_threshold_1978} could potentially reveal further insights on the coevolution of adaptive and neutral traits. For simplicity, here we explored a simpler scenario that nonetheless captures some specific aspects of subsistence strategies such as farming. 
We note that with our focus on understanding the spread of farming, the analysis of the first phase, i.e. before the population has reached its carrying capacity, is of particular interest. It is widely accepted that farmers usually migrate to new areas in small numbers and experience some sort of competitive growth advantage compared to the incumbent hunter-gatherer communities.If a hitchhiking signature has not been established in this first phase, it is  unlikely to happen in later phases. Naturally, a precise analysis of later phases of the spread dynamic may involve the inclusion of population regulation which is outside the scope of this paper. }
